---
title: "ITRDA3-11 Assignment"
output: html_notebook
---
Question One: Data Pre-processing


Question 1.a)
```{r}
library(readr)

graduate_survey <- read_csv("graduate_survey.csv")
head(graduate_survey)

```
```{r}
dim(graduate_survey)         # Check number of rows & columns
```
```{r}
colnames(graduate_survey)    # Display column names
```
```{r}
summary(graduate_survey)     # Quick summary of the data
```
```{r}
# Selecting the relevant columns 
library(dplyr)

cols_selected <- graduate_survey %>%
  select(Campus,StudyField,Branch,Role,EduLevel,ProgLang,Databases,Platform,WebFramework,Industry,AISearch,AITool,Employment)

cols_selected

# We are selecting these columns as there are over 80 and not all are relevant to the scenario.
# We are focusing on fields that directly relate to the graduates education and careers.
# Redundant or unnecessary columns are removed in order to improve efficiency.
```
Question 1.b) Missing data can cause inaccurate insights. If a significant number of values are missing, certain techniques should be used to fill or remove them.
```{r}
# Check if there are any missing values 
sum(is.na(cols_selected))
```
```{r}
# Check missing values per column
colSums(is.na(cols_selected))
```
```{r}
# Check the percentage of missing values per column
missing_percent <- colSums(is.na(cols_selected)) / nrow(cols_selected) * 100
missing_percent_rounded <- round(missing_percent, 2)  # Rounds off to 2 decimal places
print(missing_percent_rounded)


```
```{r}
 str(cols_selected)  # Check structure of data set
```
```{r}
str(cols_selected$EduLevel) # checking the format of the EduLevel column, as it might be a column with numbers and characters
```

```{r}
# Filling in the NAs with Unknown
cols_selected$EduLevel <- as.character(cols_selected$EduLevel)
cols_selected[is.na(cols_selected)] <- "Unknown" 

```

```{r}
# Filling in the missing values, cant remove them as there is to many within the data set. If we remove rows and columns it could potentially cause disruptions
cols_selected[is.na(cols_selected)] <- "Unknown"
```

Checking if there are any missing values that remain:
```{r}
colSums(is.na(cols_selected))
```
Question 1.c) Standardizing Categorical columns
```{r}
# Here I am checking for unique values across the selected columns, that might need to be standardized.

sapply(cols_selected, function(x) length(unique(x))) # Here we are checking amount of unique values per column
```
```{r}
str(cols_selected)
```
```{r}
library(dplyr)

cols_selected <- cols_selected %>%
  mutate(
    # Standardizing Campus Names
    Campus = case_when(
      Campus == "Umhlanga" ~ "Durban",
      Campus == "JHB" ~ "Johannesburg",
      Campus == "CPT" ~ "Cape Town",
      TRUE ~ Campus  # Keep other values unchanged
    ),
    
    # Standardizing Study Field
    StudyField = case_when(
      StudyField %in% c("it", "IT") ~ "Information Technology",
      StudyField == "data science" ~ "Data Science",
      TRUE ~ StudyField
    ),
    
    # Standardizing Branch
    Branch = case_when(
      Branch %in% c("Software Development", "Software Eng") ~ "Software Engineering",
      Branch == "Data Science" ~ "Data Science",
      TRUE ~ Branch
    ),
    
    # Standardizing Job Role
    Role = case_when(
      Role %in% c("Software Engineer", "Software Developer", "Dev") ~ "Software Developer",
      Role %in% c("Data Scientist", "Data Analyst") ~ "Data Professional",
      TRUE ~ Role
    ),
    
    # Standardizing Education Level
    EduLevel = case_when(
      EduLevel == "Bsc" ~ "BSc",
      EduLevel == "Msc" ~ "MSc",
      EduLevel == "phd" ~ "PhD",
      TRUE ~ EduLevel
    )
  )

# Not standardizing columns with large unique values as it is time consuming.
```

```{r}
sapply(cols_selected, function(x) length(unique(x)))  # Count unique values per column

```
Question 1.d) Subsetting
```{r}
library(dplyr)

# Count number of responses per campus
campus_counts <- cols_selected %>%
  group_by(Campus) %>%
  summarise(response_count = n()) %>%
  arrange(desc(response_count))

print(campus_counts)  # View campus response counts
```
```{r}
# Get the top 5 campuses
top_campuses <- campus_counts %>%
  top_n(5, response_count) %>%
  pull(Campus)  # Extract only the campus names

print(top_campuses)  # Check which campuses are selected

```

Now we filter the top 5 campuses with the most responses:
```{r}
campus_filtered <- cols_selected %>%
  filter(Campus %in% top_campuses)

# Check if filtering worked
table(campus_filtered$Campus)  # Confirm only top campuses remain

```

Question Two: Visualizations 

1.2) Top tools used 

Top 10 Programming Languages
```{r}
# Find the most used programming languages 
library(tidyverse)

# Step 1: Select the column and remove missing values
prog_langs <- campus_filtered %>% # we used campus filtered as it is the cleansed data containing data from the campuses with the most responses.
  select(ProgLang) %>%
  filter(!is.na(ProgLang))

# Step 2: Split the values by ";"
prog_langs <- prog_langs %>%
  separate_rows(ProgLang, sep = ";")

# Step 3: Count occurrences of each programming language
prog_lang_counts <- prog_langs %>%
  group_by(ProgLang) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

# Step 4: View the results
print(prog_lang_counts)



```
```{r}
ggplot(prog_lang_counts[1:10, ], aes(x = reorder(ProgLang, count), y = count, fill = ProgLang)) +
  geom_bar(stat = "identity") +
  coord_flip() +  # Rotates the bars for better readability
  labs(title = "Top 10 Programming Languages Used by Graduates", x = "Programming Language", y = "Count") +
  theme_minimal()

# As can be seen on the bar graph below, JavaScript is the most commonly used programming language.
```
```{r}
# Top Databases used

db_counts <- campus_filtered %>%
  select(Databases) %>%
  filter(!is.na(Databases)) %>%
  separate_rows(Databases, sep = ";") %>%
  group_by(Databases) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

ggplot(db_counts[1:10, ], aes(x = reorder(Databases, count), y = count, fill = Databases)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Databases Used by Graduates", x = "Databases", y = "Count") +
  theme_minimal()

# As can be seen from the bar graph below, PostgreSQL is the most commonly used Database by the graduates.
```
```{r}
# Top AI Search tools
aisearch_counts <- campus_filtered %>%
  select(AISearch) %>%
  filter(!is.na(AISearch)) %>%
  separate_rows(AISearch, sep = ";") %>%
  group_by(AISearch) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

ggplot(aisearch_counts[1:10, ], aes(x = reorder(AISearch, count), y = count, fill = AISearch)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 AI Search Tools Used by Graduates", x = "AI Search Tools", y = "Count") +
  theme_minimal()

```
As can be seen from the bar graph above, the most commonly used AI Search tool amongst the graduates is ChatGPT.

```{r}

# Web Frameworks
webframeworks_counts <- campus_filtered %>%
  select(WebFramework) %>%
  filter(!is.na(WebFramework)) %>%
  separate_rows(WebFramework, sep = ";") %>%
  group_by(WebFramework) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

ggplot(webframeworks_counts[1:10, ], aes(x = reorder(WebFramework, count), y = count, fill = WebFramework)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Web Frameworks Used by Graduates", x = "Web Frameworks", y = "Count") +
  theme_minimal()

```
As can be seen from the above bar graph, the most used web framework is Node.js

```{r}

# Platforms
Platform_counts <- campus_filtered %>%
  select(Platform) %>%
  filter(!is.na(Platform)) %>%
  separate_rows(Platform, sep = ";") %>%
  group_by(Platform) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

ggplot(Platform_counts[1:10, ], aes(x = reorder(Platform, count), y = count, fill = Platform)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 10 Platforms Used by Graduates", x = "Platforms", y = "Count") +
  theme_minimal()

```
Question 2.ii) Most popular industry
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Process the industry data
industry_counts <- campus_filtered %>%
  select(StudyField, Industry) %>%
  filter(!is.na(Industry)) %>%
  separate_rows(Industry, sep = ";") %>%  # Splitting multiple industries
  group_by(StudyField, Industry) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count))

# Step 2: check if the data is structured correctly
print(industry_counts)

```
```{r}
str(industry_counts)
```

```{r}
industry_counts$Industry <- abbreviate(industry_counts$Industry, minlength = 15)  # Shorten the names to fit on bar graph

ggplot(industry_counts, aes(x = fct_reorder(Industry, count), y = count, fill = StudyField)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(title = "Most Popular Industries for Each Study Field",
       x = "Industry (Abbreviated)",
       y = "Count") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotates the labels slightly
    axis.text.y = element_text(size = 10),
    legend.position = "bottom"
  )


```
Question 2.iii)
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

role_counts <- campus_filtered %>%
  select(StudyField, Role) %>%
  filter(!is.na(Role)) %>%
  separate_rows(Role, sep = ";") %>%  # Splitting multiple roles
  group_by(StudyField, Role) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count))

print(role_counts)

```

```{r}
top_roles <- role_counts %>%
  top_n(10, count)  # Selects the top 10 job roles by count

ggplot(top_roles, aes(y = fct_reorder(Role, count), x = count, fill = StudyField)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(title = "Top 10 Job Roles for Each Study Field",
       x = "Count",
       y = "Job Role") +
  coord_flip() +  
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),  
    legend.position = "right"
  )


```
Question 2.iv) 
```{r}
# We group by studyfield and calculate percentage of employed graduates. 
library(dplyr)
library(ggplot2)

# Step 1: Process employment data
employment_summary <- campus_filtered%>%
  select(StudyField, Employment) %>%
  filter(!is.na(Employment)) %>%
  mutate(Employed = ifelse(Employment %in% c("Employed", "Self-Employed"), 1, 0)) %>%  # Count employed grads
  group_by(StudyField) %>%
  summarise(
    total_grads = n(),
    employed_grads = sum(Employed),
    employment_rate = (employed_grads / total_grads) * 100
  ) %>%
  arrange(desc(employment_rate))

print(employment_summary)

# Here in this output it can be seen the data is not displaying correctly. 



```
```{r} 
unique(campus_filtered$Employment) # used to check the values within the column and as can be seen there are several employed graduate, we have to remove ";" and detect employment within the dataset. 

```




```{r}
library(dplyr)
library(tidyr)
library(stringr)

employment_summary <- campus_filtered %>%
  select(StudyField, Employment) %>%
  filter(!is.na(Employment)) %>%
  separate_rows(Employment, sep = ";") %>%  
  mutate(Employment = str_trim(Employment)) %>%  # Remove spaces
  mutate(Employment = tolower(Employment)) %>%  # Convert to lowercase for safe matching
  mutate(Employed = ifelse(grepl("employed|self-employed|independent contractor", Employment, ignore.case = TRUE), 1, 0)) %>%  # DETECT EMPLOYMENT ANYWHERE IN THE TEXT, detects patterns,searches for employment-related words ANYWHERE in the text.
  group_by(StudyField) %>%
  summarise(
    total_grads = n(),
    employed_grads = sum(Employed),
    employment_rate = (employed_grads / total_grads) * 100
  ) %>%
  arrange(desc(employment_rate))

print(employment_summary)


```
```{r}
ggplot(employment_summary, aes(x = "", y = employment_rate, fill = StudyField)) +
  geom_bar(stat = "identity", width = 1) +
  geom_text(aes(label = paste0(round(employment_rate, 1), "%")), 
            position = position_stack(vjust = 0.5),  # Centers text inside the slices
            size = 5, color = "white") +  # Adjusts text size & color
  coord_polar("y", start = 0) +
  labs(title = "Employment Rate by Study Field") +
  theme_void() +
  theme(legend.position = "right")

```
```{r}
# Save to  CSV file: 
write.csv(campus_filtered, "campus_filtered.csv", row.names = FALSE)
```

```{r}
install.packages("usethis")
usethis::use_git()
```


